---
title: "Partial Autocorrelation"
---

We have seen controlled $MA(q)$ and $AR(p)$ examples where explicit coefficients have been given.
Real world data, however, is not easy to be modelled. Partial autocorrelation function (PACF) will helps us with that.

Note that, from now and on, we will try to stick to the [Tidyverse Style Guide for R](https://style.tidyverse.org/syntax.html#assignment-1).

Before we start, let's install some of the packages needed for this notebook:

```{r}
install.packages("ppcor")
install.packages("isdals")

library(isdals)
library(ppcor)
```


# Partial Correlation

The partial correlation is the degree of association between two random variables when the effect of a set of controlling random variables has been removed.

It is useful when we want to measure the correlation coefficient between variables and there is a confounding variable which is numerically related to those variables. When this happens, the correlation coefficient is misleading.

Given variables $X$ and $Y$, partial correlation given a set of controllig variables $Z = \{Z_1, Z_2, \ ... \, Z_n\}$ is the correlation between:

- $e_X$, which is the set of residuals from the linear regression of $X$ using $Z$.
- $e_Y$, which is the set of residuals from the linear regression of $Y$ using $Z$.

Partial correlation is then computed as:

$$
\hat{\rho}_{XY · Z} = corr[e_x, \ e_Y]
$$

Where $corr$ is the conventional correlation coefficient.

## Partial correlation in R

Let's compute the partial correlation in R by using the $bodyfat$ dataset:


```{r}
data(bodyfat)
attach(bodyfat)
pairs(cbind(Fat, Triceps, Thigh, Midarm))
```

We see *Tigh* and *Triceps* can be good predictors of *Fat* but they seem to be highly correlated:


```{r}
cor(cbind(Fat, Triceps, Thigh, Midarm))
```


The correlation between *Fat* and *Triceps* after removing the linear effects caused by *Thigh* can be estimated using the *pcor* command:


```{r}
pcor(cbind(Fat, Triceps, Thigh))
```

We see the correlation between *Fat* and *Triceps* measure (i.e. ~0.1750) is low when we remove the effect of *Thigh*. We can also manually compute it:

```{r}
Fat.hat = predict(lm(Fat ~ Thigh))
Triceps.hat = predict(lm(Triceps ~ Thigh))
cor((Fat - Fat.hat), (Triceps - Triceps.hat))
```

We can include multiple variables in the controlling set:

```{r}
pcor(cbind(Fat, Triceps, Thigh, Midarm))
```

We see that, when we also include *Midarm* in the controlling set, the partial correlation of *Fat* and *Triceps* also increases. Let's replicate the result manually:

```{r}
Fat.hat = predict(lm(Fat ~ Thigh + Midarm))
Triceps.hat = predict(lm(Triceps ~ Thigh + Midarm))
cor((Fat - Fat.hat), (Triceps - Triceps.hat))
```

## Partial Autocorrelation in timeseries

Given we have a time series $X_t$, the partial autocorrelation function at lag $k$ (i.e. $\alpha(k)$), is the autocorrelation between $X_t$ and $X_{t+k}$ when the linear dependece of variables between $X_{t+1}$ and $X_{t+k-1}$ with $X_t$ has been removed. It is defined as::

$$
\Bigg\{
  \begin{array}{ll}
  \alpha(1) = corr(X_{t+1}, X_{t}) & k = 1 \\
  \alpha(k) = corr(X_{t+k} - \hat{X}_{t+k}, X_t - \hat{X}_t) & k \ge 2
  \end{array}
$$

Where:

- $\hat{X}_t$ are the predictions of $X_t$ regressing on variables from $X_{t+1}$ to $X_{t+k-1}$.
- $\hat{X}_{t+k}$ are the predictions of $X_{t+k}$ regressing on variables from $X_{t+1}$ to $X_{t+k-1}$.

The partial autocorrelation function is a useful tool for identifying the order of an autoregressive process.

## Partial autocorrelation in R

### Example 1

Let's remember, first, the general formula for autoregressive $AR(p)$:

$$
X_t = Z_t + \phi_1 X_{t-1} + \ ... \ + \phi_p X_{t-p}
$$

Let's simulate an $AR(p)$:

$$
X_t = Z_t + 0.6 X_{t-1} + 0.2 X_{t-2}
$$

Let's code it in R and plot the raw data, the ACF and the PACF:

```{r}
par(mfrow = c(3, 1))
phi.ts1.1 <- 0.6
phi.ts1.2 <- 0.2
data.ts1 <- arima.sim(n = 500, list(ar = c(phi.ts1.1, phi.ts1.2)))

main_plot1_title <- paste(
  "AR(q=2) process with phi1=",
  phi.ts1.1,
  ", phi2=",
  phi.ts1.2,
  sep = ""
)
plot(data.ts1, main = main_plot1_title)
acf(data.ts1, main = "Autocorrelation function")
acf(data.ts1, type = "partial", main = "Partial Autocorrelation Function")
```

ACF for moving averages shows $q + 1$ statistical spikes (i.e. including spike at $0$). However, the ACF of autoregressive processes is hard to interpret
and even harder to be able to extract insights about order of the process. We can see there is a relation between the significant spikes in the PACF and
the order of the AR process.

Let's do the same for the following $AR(3)$:

$$
X_t = Z_t + 0.9 X_{t-1} - 0.6 X_{t-2} + 0.3 X_{t-3}
$$

```{r}
par(mfrow = c(3, 1))
phi.ts2.1 <- 0.9
phi.ts2.2 <- -0.6
phi.ts2.3 <- 0.3
data.ts2 <- arima.sim(n = 500, list(ar = c(phi.ts2.1, phi.ts2.2, phi.ts2.3)))

main_plot2_title <- paste(
  "AR(q=3) process with phi1=",
  phi.ts2.1,
  ", phi2=",
  phi.ts2.2,
  ", phi3=",
  phi.ts2.3,
  sep = ""
)
plot(data.ts2, main = main_plot2_title)
acf(data.ts2, main = "Autocorrelation function")
acf(data.ts2, type = "partial", main = "Partial Autocorrelation Function")
```

We confirm that there are as many significant spikes in the PACF plot as the order of the AR process generating the data.

### Beveridge Wheat Price dataset

First, let's download the data:

```{r}
download.file(
  "https://people.sc.fsu.edu/~jburkardt/datasets/time_series/beveridge_wheat.txt",
  "wheat_price_dataset.txt"
)
```

Then let's load it and transform it. We are going to grab a subset of the data and create a moving average, which will be shown in red
in the plot:

```{r}
beveridge <- read.table("wheat_price_dataset.txt")
beveridge.ts = ts(beveridge[, 2], start = 1500)
plot(beveridge.ts, ylab = "price", main = "Beveridge Wheat Price data")

# Create moving average using 15 points up and down each point
beveridge.ma = filter(beveridge.ts, filter = rep(1/31, times = 31), sides = 2)
lines(beveridge.ma, col = "red")
```

Beveridge wanted to get a stationary dataset by scaling the original dataset by the moving average computed. Let's reproduce this:

```{r}
par(mfrow = c(3, 1))
beveridge.scaled_ts = beveridge.ts / beveridge.ma

plot(beveridge.scaled_ts, ylab = "Scaled price", main = "Scaled Beveridge data")

# na.omit removes the empty data at the beggining and end of the moving average

acf(na.omit(beveridge.scaled_ts),
    main = "Autocorrelation function for Scaled Beveridge data")

acf(na.omit(beveridge.scaled_ts),
    type = "partial",
    main = "Autocorrelation function for Scaled Beveridge data")
```


We see:

- Scaled data seems relatively stationary (though there may be seasonality, which does not concerns us now).
- PACF has two spikes, which may be evidence that we are dealing with a second order process.

R has a built-in function called $ar$ which provides the autoregressive $\phi$ coefficients and the order of
the underlying process given a time series and a maximum order to consider:

```{r}
ar(na.omit(beveridge.scaled_ts), order.max = 5)
```

We see the function comes up with a process of order two, which we already derived from the PACF.
