---
title: "Estimating AR(3) coefficients"
---

Given definition of $AR(3)$:

$$
X_t = Z_t + \phi_1 X_{t-1} + \phi_2 X_{t-2} + \phi_3 X_{t-3} \\
Z_t \sim Normal(\mu, \sigma_Z^2)
$$

We will be simulating the following process:

$$
X_t = Z_t + \frac{1}{3} X_{t-1} + \frac{1}{2} X_{t-2} + \frac{7}{100} X_{t-3} \\
Z_t \sim Normal(\mu, 4^2)
$$

Let's visualize the Yule Walker equations in the matrix form:

$$
\hat{b} = \hat{R} \hat{\phi} \\
\begin{bmatrix}
r_1 \\
r_2 \\
r_3 \\
\end{bmatrix} = 
\begin{bmatrix}
1 & r_1  & r_2 \\
r_1 & 1  & r_1 \\
r_2 & r_1 & 1\\
\end{bmatrix}
\begin{bmatrix}
\hat{\phi_1} \\
\hat{\phi_2} \\
\hat{\phi_3} \\
\end{bmatrix}
$$

Where $r_i$ is the estimated $i_{th}$ autocorrelation coefficient. The Yule Walker estimation for $\sigma_Z^2$ is:

$$
\hat{\sigma_Z^2} = c_0 (1 - b^{\top} \hat{\phi})
$$

Where $c_0$ is the autocovariance estimate at lag $0$.

Let's simualte the process:

```{r}
set.seed(2017)
n <- 100000
ar.process <- arima.sim(n, model = list(ar = c(1/3, 1/2, 7/100)), sd = 4)
```

Now we compute $\hat{b}$:

```{r}
r <- acf(ar.process, type = "correlation", plot = FALSE)$acf[2:4]
b <- matrix(r, 3, 1)
b
```

Now time to build $\hat{R}$:



$$
\hat{b} = \hat{R} \hat{\phi} \\
\begin{bmatrix}
r_1 \\
r_2 \\
r_3 \\
\end{bmatrix} = 
\begin{bmatrix}
1 & r_1  & r_2 \\
r_1 & 1  & r_1 \\
r_2 & r_1 & 1\\
\end{bmatrix}
\begin{bmatrix}
\hat{\phi_1} \\
\hat{\phi_2} \\
\hat{\phi_3} \\
\end{bmatrix}
$$

Let's implement a generic function for getting $\hat{R}$:

```{r}
p <- 3
R <- matrix(1, p, p)
for (row in 1:p) {
  for (i in seq(1, row - 1, length.out = row - 1)) {
    R[row, i] <- b[row - i]
  }
  for (i in seq(row + 1, p, length.out = p - row)) {
    R[row, i] <- b[i - row]
  }
}
```

We can now solve for $\hat{\phi}$:

```{r}
phi.hat <- solve(R, b)
phi.hat
```

And we can finally estimate $\sigma_Z^2$ as:

$$
\hat{\sigma_Z^2} = c_0 (1 - \hat{b}^{\top} \hat{\phi})
$$
```{r}
c0 <- acf(ar.process, type = "covariance", plot=FALSE)$acf[1]
var.hat = c0 * (1 - t(b) %*% phi.hat)
var.hat
```

We see the estimated coefficients and sigma are very close to the ones from the underlying generating process.