---
title: "Fitting recruitment data"
---

## The Data

- Recruitment (i.e. number of fish) for 453 motnhs between 1950 and 1987.
- Monthly time series.

Dataset related to this book [see @timeseriesnalaysis].

```{r}
library(astsa)
plot(rec, main = "Recruitment time series", col = "blue", lwd = 3)
```

We see the mean is constant, though we also observe some seasionality (which we are going to ignore here). Let's take a look at the ACF and PACF:

```{r}
par(mfrow=c(2, 1))
acf(rec, type = "correlation", main = "Autocorrelation function for recruitment data")
pacf(rec, main = "Autocovariance function for recruitment data")
```

We observe the PACF function has two significant peaks, which may indicate that the underlying process in an
autoregressive process of order 2. sing the *parsimony principle*, we will select the simplest of the compellig 
theories (i.e. we only have one, at this point) and try to find evidence to support it.

So we are assuming our data comes from the following autoregressive process:

$$
X_t = Z_t + \hat{\phi}_0 + \hat{\phi}_1 X_{t-1} + \hat{\phi}_2 X_{t-2} \\
Z_t \sim Normal(0, \sigma_Z^2)
$$

## Modelling the data

Let's first center data around 0 and compute the sample autocorrelation vector $r$ (i.e. $\hat{b}$):

```{r}
p <- 2
rec.centered <- rec - mean(rec)
r <- acf(rec.centered, type = "correlation", plot=FALSE)$acf[2:(2+p-1)]
r
```


Let's generate matrix $\hat{R}$:

```{r}
R.hat <- matrix(1, p, p)
for (row in 1:p) {
  for (col in 1:p) {
    if (row != col){
      R.hat[row, col] <- r[abs(row - col)]
    }
  }
}
R.hat
```

Then we can solve:

$$
\hat{R} \hat{\phi} = \hat{b}
$$

```{r}
b.hat <- matrix(r, p, 1)
phi.hat <- solve(R.hat, b.hat)
phi.hat
```

Let's obtain now the estimate of the variance of $Z_t$:

```{r}
c0 <- acf(rec.centered, type = "covariance", plot=FALSE)$acf[1]
var.hat <- c0 * (1 - t(b.hat) %*% phi.hat)
var.hat
```

We have computed the $\phi$ coefficients for the following model:

$$
X_t - \mu  = Z_t + \hat{\phi_1} (X_{t-1} - \mu) + \hat{\phi_2} (X_{t-2} - \mu) + \ldots + \hat{\phi_p} (X_{t-p} - \mu) \\
X_t = \mu + Z_t + \hat{\phi_1} (X_{t-1} - \mu) + \hat{\phi_2} (X_{t-2} - \mu) + \ldots + \hat{\phi_p} (X_{t-p} - \mu) \\
X_t = \mu + Z_t + \sum^{p}_{i=1} \hat{\phi_i} (X_{t-i} - \mu) \\
X_t = \mu + Z_t + \sum^{p}_{i=1} \hat{\phi_i} X_{t-i} - \mu \sum^{p}_{i=1} \hat{\phi_i} \\
$$

But we want to obtain the formula for the original time series:


$$
X_t = \hat{\phi}_0 + Z_t + \hat{\phi_1} X_{t-1} + \hat{\phi_2} X_{t-2} + \ldots + \hat{\phi_p} X_{t-p} \\
X_t = \hat{\phi}_0 + Z_t + \sum^{p}_{i=1} \hat{\phi_i} X_{t-i}
$$

If we equal both right-hand side expressions, we get the equation to obtain $\phi_0$:

$$
\hat{\phi}_0 + Z_t + \sum^{p}_{i=1} \hat{\phi_i} X_{t-i} = \mu + Z_t + \sum^{p}_{i=1} \hat{\phi_i} X_{t-i} - \mu \sum^{p}_{i=1} \hat{\phi_i}\\
\hat{\phi}_0 = \mu - \mu \sum^{p}_{i=1} \hat{\phi_i} \\
\hat{\phi}_0 = \mu \left(1 - \sum^{p}_{i=1} \hat{\phi_i} \right)
$$

Let's compute it in $R$:

```{r}
theta.0 <- mean(rec) * (1 - sum(phi.hat))
theta.0
```

We can then obtain the complete form of the model for the data:

$$
X_t = Z_t + 7.0333 + 1.3316 X_{t-1} - 0.4445 X_{t-2} \\
Z_t \sim (0, 94.1713)
$$

## Bibliography

---
references:
- id: timeseriesnalaysis
  title: Time Series Analysis and Its Applications
  author:
  - family: Suhmway
    given: Robert H.
  - family: Stoffer
    given: David S.
  publisher: Springer-Verlag
  isbn: 0387989501
  type: book
  issued:
    year: 2005
---
