---
title: "Fitting Johnson & Johnson data  "
---

## The Data

- Earnings per Johnson & Johnson share in dollars between 1960 and 1980.
- Quarter data.

Dataset related to this book [see @timeseriesnalaysis].

```{r}
library(astsa)
plot(
  JohnsonJohnson,
  main = "Johnson & Johnson quarterly earnings",
  col = "blue",
  lwd = 3,
  ylab = "$"
)
```

Data shows an increasing trend but also a systematic increase of the variance. In order to have constant mean, we can use the **log return** $r$ as:

$$
r_t = log \left( \frac{X_t}{X_{t-1}} \right) = log(X_t) - log(X_{t-1})
$$

Let's do it in R:


```{r}
jj <- JohnsonJohnson
jj.centered <- diff(log(jj))

par(mfrow = c(3, 1))
plot(
  jj.centered,
  main = "Johnson & Johnson quarterly log return earnings",
  col = "blue",
  lwd = 3,
  ylab = "$"
)
acf(
  jj.centered,
  type = "correlation",
  main = "Autocorrelation function for J&J log return data"
)
pacf(
  jj.centered,
  main = "Autocovariance function for recruitment J&J log return data"
)
```

We see it now has an approximately constant mean. We see PACF has 4 significant peaks, so, following the same principle as in the
previous example, we are going to try for the simplest and most compelling theory, which is that the underlying process
is an autoregressive process of order 4:

$$
r_t = Z_t + \hat{\phi}_0 + \hat{\phi}_1 r_{t-1} + \hat{\phi}_2 r_{t-2} + \hat{\phi_3} r_{t-3} + \hat{\phi_4} r_{t-4} \\
Z_t \sim Normal(0, \sigma_Z^2)
$$

## Modelling the data

Let's first center data around 0 and compute the sample autocorrelation vector $r$ (i.e. $\hat{b}$):

```{r}
p <- 4
jj.centered.zero_mean <- jj.centered - mean(jj.centered)
r <- acf(jj.centered.zero_mean, type = "correlation", plot=FALSE)$acf[2:(2+p-1)]
r
```


Let's generate matrix $\hat{R}$:

```{r}
R.hat <- matrix(1, p, p)
for (row in 1:p) {
  for (col in 1:p) {
    if (row != col){
      R.hat[row, col] <- r[abs(row - col)]
    }
  }
}
R.hat
```

Then we can solve:

$$
\hat{R} \hat{\phi} = \hat{b}
$$

```{r}
b.hat <- matrix(r, p, 1)
phi.hat <- solve(R.hat, b.hat)
phi.hat
```

Let's obtain now the estimate of the variance of $Z_t$:

```{r}
c0 <- acf(jj.centered.zero_mean, type = "covariance", plot=FALSE)$acf[1]
var.hat <- c0 * (1 - t(b.hat) %*% phi.hat)
var.hat
```

We can finally estimate $\phi_0$ using the formula derived from the previous example:

$$
\hat{\phi}_0 = \mu \left(1 - \sum^{p}_{i=1} \hat{\phi_i} \right)
$$

Let's compute it in $R$:

```{r}
theta.0 <- mean(jj.centered) * (1 - sum(phi.hat))
theta.0
```

We can then obtain the complete form of the model for the data:

$$
r_t = Z_t + 0.0798 - 0.6293 \ r_{t-1} - 0.5171 \ r_{t-2} - 0.4883 \ r_{t-3} + 0.2651 \ r_{t-4} \\
Z_t \sim Normal(0, 0.0142)
$$


## Bibliography

---
references:
- id: timeseriesnalaysis
  title: Time Series Analysis and Its Applications
  author:
  - family: Suhmway
    given: Robert H.
  - family: Stoffer
    given: David S.
  publisher: Springer-Verlag
  isbn: 0387989501
  type: book
  issued:
    year: 2005
---
