---
title: "Forecast with simple exponential smoothing"
---

## Data

Total anual rainfall in London in inches, between 1813 and 1912 (Original data from Hipe and McLeod, 1994).

```{r}
rain.data <- scan(
  "https://robjhyndman.com/tsdldata/hurst/precip1.dat", skip = 1
)
rain.ts <- ts(rain.data, start = c(1813))
```

Let's visualize the data:

```{r}
par(mfrow = c(1, 2))
hist(
  rain.data,
  main = "Annual London Rainfall 1813-1912",
  xlab = "Rainfall in inches"
)
qqnorm(rain.data, main = "Q-Q plot of London Rainfall 1813-1912")
qqline(rain.data)
```

There seems to be a structural deviation from normality, as the Q-Q plot shows, though it is not very extreme.

```{r}
par(mfrow = c(2, 1))
plot(rain.ts, main = "Annual London Rainfall 1813-1912")
acf(rain.ts, main = "ACF of London Rainfall 1813-1912")
```

Data seems to be quite random, though some seasonality may be appreciated. What if we try to fit an arima model?

```{r}
library(forecast)
auto.arima(rain.ts)
```

The `auto.arima` command says the model $X_t = 24.8239$ is the best approach it could find. However, could we model it in any other way?

## Simple exponential smoothing (SES)

Given $X^t_{t+k}$ is the forecast of future point $k$ timesteps away using data up to time $t$, we define a Simple Exponential Smoothing (SES) as forecast method that uses a weighted average of the previous values, so recent values have higher weights than those which are earlier in time.

In order to select a decaying sequence of weights for successive timesteps, we go again into geometric series, as they decay at a constant ratio. Let's see the geometric series with $a=1$ and $r=(1-\alpha)$:

$$
\sum^{\infty}_{k = 0} (1 - \alpha)^k = 1 + (1 - \alpha)  + (1 - \alpha)^2 + \ldots = \frac{1}{1 - (1 - \alpha)} = \frac{1}{\alpha}
$$

In order to make the geometric series sum 1, we can multiply by $\alpha$ at both sides:

$$
\alpha \sum^{\infty}_{k} (1 - \alpha)^k = \alpha + \alpha(1 - \alpha)  + \alpha(1 - \alpha)^2 + \ldots = \alpha \frac{1}{1 - (1 - \alpha)} = \frac{\alpha}{\alpha} = 1
$$

We can forecast next value, using Simple Exponential Smoothing, by weighting previous values using the corresponding term in the geometric series above:

$$
X^t_{t+1} = \alpha X_t{} + \alpha(1 - \alpha) X_{t-1} + \alpha(1 - \alpha)^2 X_{t-2} + \ldots
$$
Given that:

$$
X^{t-1}_{t} = \alpha X_{t-1} + \alpha (1 - \alpha) X_{t-2} + \alpha (1 - \alpha)^2 X_{t-3} + \ldots
$$

Then we can write it as a recurrent expression:

$$
X^t_{t+1} = \begin{cases}
X_1 & t = 1 \\
X^t_{t+1} = \alpha X_t + (1 - \alpha) X^{t-1}_t & t \gt 1
\end{cases}
$$

Note that the $\alpha$ value controls the behavior of the prediction:

- Large values of $\alpha$ (i.e. $\alpha \approx 1$) put emphasis on *near* observations. 
- Smaller values of $\alpha$ allow weighting data points that are further in the past.


### SES in R

We can implement Smooth Exponential Smoothing easily using R:

```{r}
alpha <- 0.2
forecast.values <- NULL
n <- length(rain.data)

# First value is a naive forecast
forecast.values[1] <- rain.data[1]

for (i in 1:n) {
  forecast.values[i + 1] <- alpha * rain.data[i] + (1 - alpha) * forecast.values[i]
}

paste("Forecast for time", n + 1, "=", forecast.values[n + 1])
```

Note we are using an arbitrary value of $\alpha$. In order to find the best $\alpha$ value, we can select the one that produces the minimal Sum of Squares Error (SSE):

```{r}
sses <- NULL
forecast.values <- NULL
alpha.values <- seq(0.001, 0.999, by = 0.001)

for (k in 1:length(alpha.values)) {
  alpha <- alpha.values[k]
  forecast.values[1] <- rain.data[1]

  for (i in 2:n) {
    forecast.values[i] <- alpha * rain.data[i - 1] + (1 - alpha) * forecast.values[i - 1]
  }
  
  sses[k] <- sum((rain.data - forecast.values)**2)
}

plot(sses ~ alpha.values, main = "SEE for different alpha values")

smallest.sses.idx = which.min(sses)
print(paste("Optimal alpha is", alpha.values[smallest.sses.idx]))

```

### Built-in R function


```{r}
HoltWinters(rain.ts, beta = FALSE, gamma = FALSE)
```

Note value is slightly different as this method is using a finer grid than the one in the example above.
